White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-11-14 papers/wdmerger_II/paper.tex
Rework the suppressed burn section

We terminate the simulation at $t = 9$ seconds, as we have found that for this
particular setup, all of the significant burning has completed by this time. At
this point, the total energy on the domain has become positive, typically about
$10^{51}$ ergs, and has stopped increasing. This means that the system has become
unbound due to nuclear energy release, and that no further meaningful nuclear energy
generation is occurring. In our discussions below, it helps to have some other
anchoring points in time when significant events occur. We (arbitrarily) define the
collision time, $t_c$, as the first time when the maximum density on the grid reaches
$5 \times 10^6\ \text{g / cm}^3$. This is slightly larger than the initial maximum
density of the two WDs, and is reached only when the WDs meet at the center of the
domain. We define the detonation time, $t_d$, as the time when the temperature
first reaches $5 \times 10^9\ \text{K}$.

@@

Now let us examine the results for a collision of two 0.64 $\msolar$ WDs. We use
128 zones in the radial dimension, corresponding to a spatial resolution of 312.5 km.
Timesteps are limited only by the hydrodynamic stability constraint (and the plotfile
interval constraint). Resolutions in the range of a few hundred km have been a common
choice in earlier papers on white dwarf mergers and collisions. No adaptive mesh
refinement is used. For this configuration, the collision time is $t_c = 6.47\ \text{s}$,
and the detonation time is $t_d = 7.43\ \text{s}$. The peak density at the detonation time
is $2.3 \times 10^7\ \text{g / cm}^3$.

@@

Except for the discussion later about the suppressed burn mode, the simulations
shown in this work all use the self-heating mode. The hydrostatic and hybrid burns
are only a slight variation on the self-heating burn in practice for this problem.
The reason that the hydrostatic burn cannot substantially change the outcome is
that even though the temperature is fixed during the burn, it is \textit{not}
constant in the rest of the hydrodynamics step. And in the hydrodynamics step,
the \castro\ algorithm performs multiple EOS calls to ensure that the temperature
is synchronized with the internal energy at various points in the update. So in
practice there is a temperature change due to the burn for our problem. This seems
inevitable for any hydro code unless the hydrodynamics update includes an explicit
equation for $T$ that is independent of what is happening for the internal energy.
The hydrostatic burn does slightly limit the energy release during any given burn,
but summed up over the whole simulation this is not enough to qualitatively affect
the outcome. The hybrid burn does not affect the simulation outcome at all relative
to the hydrostatic burn. The difference between them only applies for zones that have
a net negative energy release from the hydrostatic burn, which occurs only after
we have burned to NSE. The hybrid burn was mainly implemented by \citet{raskin:2010}
for making it easier to numerically complete a burn, rather than because of the
effects it has on the outcome of the simulation.

@@

\citet{kushnir:2013} point out that an inappropriate timestep is 
not the only way for the numerical discretization to cause 
severe errors in the burning. Another failure mode is when
the energy injection timescale
$\tau_e$ is shorter than the sound-crossing time $\tau_s$ in a zone.
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is obviously a problem inherent to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
is spurious in this case and should be avoided if possible.
The goal is to ensure that the following condition holds:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $\tau_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{s}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{s} = 0.1$ 
for their simulations, and we do too (this parameter can be set 
at runtime in \castro).

\citet{kushnir:2013} implemented this criterion by artificially 
limiting the magnitude of the energy release after a burn. We
too have developed an option for our burner to do this,
the ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied. To achieve this we directly multiply the
right-hand-side vector in the integration by a constant factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $\tau_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $\tau_e$ to be the value of the
internal energy of the zone at the beginning of the burn. If
instead one allowed $c_s$ and $e$ to evolve with the burn, one
would obtain a less conservative limiter in the case of explosive
burning, as $c_s$ and $e$ are both increasing in this case.
As the point of the limiter is to ensure that the changes to the
\textit{original} energy are small enough so that the following
hydrodynamics update can advect away newly generated energy
quickly enough to avoid a numerically seeded detonation,
we desire the most conservative version of the limiter. We discuss
our results with the suppressed burning mode in \autoref{sec:collisionburning}.
Note that we have found that with this option enabled, it typically takes
many more timesteps to complete a burn than in self-heating mode.

Regardless of whether the suppressed burning mode works for this
particular problem, it is not physical, so we include for
consideration a different approach. If we insist that we
cannot directly control the energy injection timescale, we 
must find a way to alter the sound-crossing timescale. 
We can achieve this by adding levels of refinement in 
regions that do not satisfy \autoref{eq:burning_limiter_2},
which effectively lowers $\Delta x$ and thus the
sound-crossing time. We keep tagging zones for refinement
based on this criterion until the criterion is satisfied
on the finest level. Since the concern is regions that 
may detonate, we also tag nearby zones in a buffer region
which do not themselves satisfy the criterion,
so that a detonation in a single timestep cannot 
escape into non-refined regions. The width of the buffer 
region should thus be at least as large as the number of 
timesteps before a regridding procedure is performed.
We choose a value of two for both the number of zones in the 
buffer region and the number of steps in between regrids,
for all simulations in this paper.

We agree with \citet{kushnir:2013} that solving this numerical
instability is crucial to avoiding unphysical detonations.
A simulation that does not solve this problem will not obtain
the correct amount and will not converge properly with resolution.
This may justify the addition of many AMR levels to the domain
if a correct evaluation of the burning phase is desired.
Even if one cannot afford the full resolution required by this
AMR criterion, and chooses to limit the number of levels to some
predetermined maximum based on a constraint of computing time,
the added resolution will still go to the most-needed places.

A potential limitation of this approach is that it does not turn on
until \textit{after} an unstable burning region has been generated,
because we generally only perform a regridding and level generation
step at times when AMR levels are synchronized, such as at the beginning
of the timestep on the coarsest level. So this AMR criterion cannot
help in the case where a spurious detonation begins in a single timestep
on the coarse grid, though it will kick in immediately after that
timestep to resolve the regions where the detonation has occurred.
This situation is common in our white dwarf collision simulations.
We have developed a remedy to this issue. At the \textit{end} of every step
on every level, we check whether the stability criterion has been violated.
If so, we recognize that it is not too late to solve the problem -- we can
still add levels of refinement in those places. We then perform the new level
generation to cover specifically those places that violated the stability criterion.
To ensure consistency with our algorithm, we use the \textit{old}-time values
of the coarse fluid state to interpolate into the new fine level, even
though it is the \textit{new}-time values that were used to guide the level
generation. Then our algorithm evolves the timestep on the new level as normal.
Because our algorithm is recursive, this can continue to generate
as many new levels as are needed to satisfy the stability criterion.
The upshot is that as long as the simulation can afford enough levels to
satisfy the stability criterion, there never needs to be even a single timestep
that violates the criterion.

%==========================================================================
% Standard Results
%==========================================================================
\section{Collision Burning}
\label{sec:collisionburning}

% Discuss the results for the case of the uniform grid with 256 zones along
% the $z$ axis, yielding a spatial resolution of 312.5 km, with no timestep limiting.
% Discuss the effect of the burning mode here.

In \autoref{sec:burner} we observed that there is an alternative to the traditional
self-heating approach in a burn, the ``suppressed'' burning mode. This alternative
is relevant to a collision problem because, as pointed out by \cite{raskin:2010}
and \cite{kushnir:2013}, it is possible for a typical zone in our simulation to
release a very large amount of energy in a burn before it is cooled
off by a subsequent hydrodynamic expansion step, leading to the numerically unstable burning
problem described in \autoref{sec:unstable_burning}. However, the suppressed burning mode
(modeled after \citet{kushnir:2013}) produces only about
$2 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed. In order to ensure that the
very different results we obtained were not a result of the specific choice we made for how to suppress
the burn, we tried varying the approach discussed in \autoref{sec:burner}. First, we limited based on
the current value of the internal energy and sound speed rather than the fixed, initial value during a burn
(this should be a less severe limiter). This did not yield a significant difference in the outcome. Second,
we also tried applying the limiting after the burn had completed, using the differences between the input
and output states of the integration, rather than directly limiting the RHS in the ODE. The behavior of the
burn was qualitatively different, in that nickel production began much earlier in the collision when the limiting
was applied this way. Nevertheless, the qualitative final outcome of the collision -- only trace amounts
of nickel produced in total -- was still unchanged. With this variant too, it did not matter whether the limiting
factor was determined based on the initial value of the state or the final value of the state.
Our default resolution is significantly lower than that of \citeauthor{kushnir:2013} (the text
says they used resolutions up to a few km) but we have tried runs at higher resolution and
the results are qualitatively similar. We did not directly compare the results of the method at
the same resolution as theirs, but our results indicate at least that this limiter significantly
modifies the evolution for modest resolution simulations. The limiter does delay the time at which
the detonation occurs, as expected by \citeauthor{kushnir:2013}, but the conditions for generating
significant amounts of iron-group elements do not develop. It is not clear what the relevant difference
is between our simulation and that of \citeauthor{kushnir:2013}, but as the suppressed burn limiting
used here is ultimately unphysical, we choose not to use it in any other simulation.

@@

To answer these questions we performed a series of simulations at varying base
resolutions, augmented with additional runs with AMR. In each case we ran the
simulation to the point at which unstable burning was approached. Then we stopped
the run and created a checkpoint. Starting from that checkpoint, we did a simulation
at varying levels of refinement based on the unstable burning criterion. We will refer
to the refinement through the refinement factor, $r$. This factor is the jump in
resolution between the coarse grid and the finest grid. The uniform base grid case is $r = 1$.
Meanwhile, $r = 16$ means a uniform coarse grid plus two levels of 4x refinement.

>>>>

We terminate the simulation at $t = 9$ seconds, as we have found that for this
particular setup, all of the significant burning has completed by this time. At
this point, the total energy on the domain has become positive, typically about
$10^{51}$ ergs, and has stopped increasing. This means that the system has become
unbound due to nuclear energy release, and that no further meaningful nuclear energy
generation is occurring. In our discussions below, it helps to have some other
anchoring points in time when significant events occur. We (arbitrarily) define the
collision time, $t_c$, as the first time when the maximum density on the grid reaches
$5 \times 10^6\ \text{g / cm}^3$. This is slightly larger than the initial maximum
density of the two WDs, and is reached only when the WDs meet at the center of the
domain. We define the detonation time, $t_d$, as the time when the temperature
first reaches $4 \times 10^9\ \text{K}$.

@@

Now let us examine the results for a collision of two 0.64 $\msolar$ WDs. We use
128 zones in the radial dimension, corresponding to a spatial resolution of 312.5 km.
Timesteps are limited only by the hydrodynamic stability constraint (and the plotfile
interval constraint). Resolutions in the range of a few hundred km have been a common
choice in earlier papers on white dwarf mergers and collisions. No adaptive mesh
refinement is used. For this configuration, the collision time is $t_c = 6.47\ \text{s}$,
and the detonation time is $t_d = 7.43\ \text{s}$. The peak density at the detonation time
is $2.3 \times 10^7\ \text{g / cm}^3$. The detonation occurs at the center of the domain
where the two WDs initially meet.

@@

The simulations shown in this work all use the self-heating mode. The hydrostatic
and hybrid burns are only a slight variation on the self-heating burn in practice
for this problem.
The reason that the hydrostatic burn cannot substantially change the outcome is
that even though the temperature is fixed during the burn, it is \textit{not}
constant in the rest of the hydrodynamics step. And in the hydrodynamics step,
the \castro\ algorithm performs multiple EOS calls to ensure that the temperature
is synchronized with the internal energy at various points in the update. So in
practice there is a temperature change due to the burn for our problem. This seems
inevitable for any hydro code unless the hydrodynamics update includes an explicit
equation for $T$ that is independent of what is happening for the internal energy.
The hydrostatic burn does slightly limit the energy release during any given burn,
but summed up over the whole simulation this is not enough to qualitatively affect
the outcome. The hybrid burn does not affect the simulation outcome at all relative
to the hydrostatic burn. The difference between them only applies for zones that have
a net negative energy release from the hydrostatic burn, which occurs only after
we have burned to NSE. The hybrid burn was mainly implemented by \citet{raskin:2010}
for making it easier to numerically complete a burn, rather than because of the
effects it has on the outcome of the simulation. The suppressed burn has separate
issues which we defer to \autoref{sec:unstable_burning}

@@

\citet{kushnir:2013} point out that an inappropriate timestep is 
not the only way for the numerical discretization to cause 
severe errors in the burning. Another possible failure mode is when
the energy injection timescale
$\tau_e$ is shorter than the sound-crossing time $\tau_s$ in a zone.
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is of course a problem inherent only to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
may be spurious in this case. If the following condition holds,
we can be confident that a numerically seeded detonation has not occurred:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $\tau_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{s}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{s} = 0.1$ 
for their simulations. \citet{kushnir:2013} implemented this
criterion by artificially limiting the magnitude of the energy
release after a burn.

To understand this choice, consider the outcome of our baseline case.
When the two WDs collide, there is a prompt detonation
at the contact point at the center of the domain. This detonation occurs
when the density is relatively low because the WDs have not had much
time to collide, and so the nucleosynthetic yield of nickel is low. If
this is the case, the colliding WDs are not a good candidate for
normal SNe Ia. However, if this detonation is delayed, material builds
up at a stalled shock in the center, and the eventual detonation occurs
on the outer edge of the stalled shock. There will then be much more
high density material to process into nickel, and the resulting
explosion creates an amount of nickel potentially sufficient to explain
normal SNe Ia. The discussion thus centers on whether the prompt detonation
is physical. \citeauthor{kushnir:2013} argues that it is not, on the basis
that $\tau_s > \tau_e$ when this detonation occurs. We find the same thing:
for the baseline case, the sound-crossing time is about five times longer
than the energy injection time. This means that a detonation is guaranteed
to occur for numerical reasons. \citeauthor{kushnir:2013} thus argue that
the suppressed burn is justifiable, as it suppresses this prompt detonation.

This argument is flawed because a physical detonation may \textit{also}
occur this way. For example, consider a region of WD material at uniformly
high temperature, say $5 \times 10^9\ \text{K}$, with an arbitrarily large size,
say a cube with side length 100 km. This region will very likely detonate,
even if it is surrounded by much cooler material. By the time the material on
the edges can advect heat away, the material in the center will have long since
started burning carbon, as the sound crossing time scale is sufficiently large
compared to the energy injection time scale. This is true regardless of whether
the size of this cube corresponds to the spatial resolution in a simulation.
Suppression of the burn in this case is unphysical: if we have a zone matching
these characteristics, the zone should detonate.

It is true that when our resolution is low enough, there is a floor on the size
of a hotspot, possibly making such a detonation more likely. However, this is an
unavoidable consequence of the low resolution: it is the correct result of the
simulation that was performed. If the results do not match what occurs at higher
resolution, then the simulation is not converged and the results are not reliable.
However, it may also be the case that a higher resolution simulation will yield
similar results, for example because even at the higher resolution, the physical
size of the hotspot stays the same. One cannot appeal to the numerical instability
criterion alone to make a decision in this context.

Furthermore, we find that \textit{all} of the detonations we have seen at low
and moderate resolution inevitably violate the stability criterion. This is true
even if the prompt detonation is suppressed: the delayed detonation also is
numerically unstable. For comparison purposes, we developed an option for our
burner to use a ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied\footnote{To achieve this we directly multiply the
right-hand-side vector in the integration by a factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $\tau_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $\tau_e$ to be the value of the
internal energy of the zone at the beginning of the burn.}. When we
evolve the baseline case with this suppressed burner, the prompt detonation
is indeed suppressed. A later detonation does occur, but this delayed
detonation also triggers the suppression. The result is that the detonation
does not get to very high temperatures, and the result is again a ``failed''
supernova that does not generate nearly enough nickel. Thus we do not
reproduce the claim of \citeauthor{kushnir:2013} that the suppressed
burning mode suppresses only the prompt detonation, but allows the
delayed detonation to occur. We could find some way to turn off the
suppression and allow the delayed detonation, but without some independent
confirmation that the delayed detonation is physical (while the prompt
detonation is not) this would be begging the question.

The upshot of this discussion is that the only way to resolve the
question is with higher resolution simulations. The semi-analytic
criterion provided by \cite{garg:2017} indicates that typical critical
radii for detonation of equal C/O material are in the range 1-10 km.
So, as pointed out by \citeauthor{kushnir:2013}, this numerical
stability issue is mainly a concern for simulations with
resolution of O(10 km) or worse. Detonations observed at higher
resolution are physically plausible because the critical temperature
gradient will be resolved. We have developed an AMR criterion
to help achieve this resolution: we tag on regions suggested by
the burning stability criterion, \autoref{eq:burning_limiter_2}.
The original motivation for this choice was that by lowering
$\Delta x$ and thus the sound-crossing time, we could reach
the point where the detonation was occurring in a numerically stable
region. This goal was not realized due to the reasons discussed above.
Instead, we simply use it as a guide for where to refine. We tag
for refinement all zones with $10^{-6} \leq f_{s} \leq 1$, where
$f_{s}$ is the ratio of the sound crossing time to the energy injection
time. We do not target zones for refinement with $f_{s} > 1$ since
at that point a detonation is locked in. We refine up to a specified
maximum resolution ratio relative to the coarse grid.

@@

To answer these questions we performed a series of simulations at varying base
resolutions, augmented with additional runs with AMR. In each case we ran the
simulation to the point at which unstable burning was approached:
$\tau_{s} / \tau_{e} = 10^{-6}$. Then we stopped
the run and created a checkpoint. Starting from that checkpoint, we did a simulation
at varying levels of refinement based on the unstable burning criterion. We will refer
to the refinement through the refinement factor, $r$. This factor is the jump in
resolution between the coarse grid and the finest grid. The uniform base grid case is $r = 1$.
Meanwhile, $r = 16$ means a uniform coarse grid plus two levels of 4x refinement.
